### 2019-03-06 paper/sex_selection_spacing-ver2.tex
Minor edits for grammar/clarity in Introduction

% [Why spacing?] 
There are four main reasons for examining birth spacing and how it changes with 
access to sex selection.
First, researchers have made extensive use of birth spacing as a measure of son preference, 
and it is critical to understand to what extent spacing is still useful as a measure of son 
preference and develop new methods if needed.
In India, the sex of the last-born child significantly impacts the duration to next birth,
with the expected birth interval about three percent longer if the last-born was a son than
if it was a daughter \citep{Bhalotra2008,Kumar2016}.
Similarly, ethnic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.%
\footnote{
Ethnic Indians in Malaysia do, however, show little evidence of son preference, although
the sample sizes are small \citep{Pong1994}.
}
In both Bangladesh and Vietnam the more boys a family has the longer the expected duration 
to the next birth \citep{Haughton1995,Haughton1996,Rahman1993,Soest2018}.
Outside Asia, the evidence is more mixed.
North Africa shows shorter spacing in the absence of sons, while a similar pattern does
not exist in Sub-Saharan Africa \citep{Rossi2015}.%
\footnote{
A possible exception is Senegal, where women with a higher risk of widowhood show
shorter birth spacing until they have secured a son \citep{Lambert2016}.
}

Second, birth spacing can have a substantial impact on children's outcomes.
Most of the research finds a negative impact of short spacing on child health, although 
identifying the causal mechanisms have proven more difficult
\citep{Conde-Agudelo2006,Conde-Agudelo2012}.
There are three pathways that birth intervals can affect health.
First, there is maternal depletion, where the mother does not have sufficient time to 
recover after her last pregnancy before she becomes pregnant again.%
\footnote{
This depletion can take the form of nutrition, folate, or cervical insufficiency.
}
Second, disease transmission both from mother to child and between children close in age.
Third, sibling competition, where siblings close in age compete for scarce resources and
parental care.
Which mechanism dominates is of particular interest when son preference drives spacing 
decisions.
Both the maternal depletion and the transmission of disease explanations predominately
affect the subsequent child and the mother.
Only sibling competition has the potential to negatively affect the older child.

In India, there is evidence that both the prior and the subsequent child have worse health 
outcomes and a higher likelihood of dying the shorter the interval between them, although 
the effect is not symmetrical
\citep{Whitworth2002,Bhargava2003,Rutstein2005,Maitra2008,Makepeace2008,Ghosh2018}.%
\footnote{
\cite{Bhalotra2008}, however, find that birth spacing only explains a limited amount
of neonatal mortality.
}
The increased mortality risk for prior children may come from shorter breastfeeding of 
girls to ensure that the mother can conceive again quicker, and with poor water
quality this shorter breastfeeding may result both in higher risk of dying and shorter
stature if the child survives \citep{Jayachandran2011,Jayachandran2017a}.
In other countries, focus has been mostly on the subsequent birth, but there is evidence 
in countries as diverse as Bangladesh, Brazil, El Salvadore, and Pakistan that
short intervals between births have adverse health effects
\citep{Cleland1984,Curtis1993,Davanzo2008,Gribble2009,Saha2013}.

The potential effects of spacing are not restricted to early outcomes, although
there is less evidence on longer-term effects, and most comes from Western countries.
In the US, for example, a longer spacing between births increases the older sibling's
test scores on the Peabody Individual Achievement Test, although there is no
statistically significant effect on the younger sibling \citep{Buckles2012}.
In line with this, close spacing appears to increase the risk of dropping out of
high school and lower the probability of attending a post-secondary school in both the
US and Sweden \citep{Powell1993,Pettersson-Lidbom2009}.
These results are, however, not uniformly supported, with recent results from Sweden 
showing no relationship between birth spacing, even if very short, and outcomes such as 
years of education completed, earnings, and unemployment \citep{Barclay2017}.

@@

\section{Results\label{sec:results}}

>>>>

% [Why spacing?] 
There are four main reasons for examining birth spacing and how it changes with 
access to sex selection.
First, researchers have made extensive use of birth spacing as a measure of son preference, 
and it is critical to understand to what extent spacing is still useful as a measure of son 
preference.
In India, the sex of the last-born child significantly impacts the duration to next birth,
with the expected birth interval about three percent longer if the last-born was a son than
if it was a daughter \citep{Bhalotra2008,Kumar2016}.
Similarly, ethnic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.%
\footnote{
Ethnic Indians in Malaysia do, however, show little evidence of son preference, although
the sample sizes are small \citep{Pong1994}.
}
In both Bangladesh and Vietnam the more boys a family has the longer the expected duration 
to the next birth \citep{Haughton1995,Haughton1996,Rahman1993,Soest2018}.
Outside Asia, the evidence is more mixed.
North Africa shows shorter spacing in the absence of sons, while a similar pattern does
not exist in Sub-Saharan Africa \citep{Rossi2015}.%
\footnote{
A possible exception is Senegal, where women with a higher risk of widowhood show
shorter birth spacing until they have secured a son \citep{Lambert2016}.
}

Second, birth spacing can have a substantial impact on children's outcomes.
Most of the research finds a negative impact of short spacing on child health, although 
identifying the causal mechanisms have proven more difficult
\citep{Conde-Agudelo2006,Conde-Agudelo2012}.
There are three pathways through which birth intervals can affect health.
First, there is maternal depletion, where the mother does not have sufficient time to 
recover after her last pregnancy before she becomes pregnant again.%
\footnote{
This depletion can take the form of nutrition, folate, or cervical insufficiency.
}
Second, disease transmission either from mother to child or between children.
Third, sibling competition, where siblings close in age compete for scarce resources and
parental care.
Which mechanism dominates is of particular interest when son preference drives spacing 
decisions.
Both the maternal depletion and the transmission of disease explanations predominately
affect the subsequent child and the mother.
Only sibling competition has the potential to affect the older child negatively.

In India, there is evidence that both the prior and the subsequent child have worse health 
outcomes and a higher likelihood of dying the shorter the interval between them, although 
the effect is not symmetrical
\citep{Whitworth2002,Bhargava2003,Rutstein2005,Maitra2008,Makepeace2008,Ghosh2018}.%
\footnote{
\cite{Bhalotra2008}, however, find that birth spacing only explains a limited amount
of neonatal mortality.
}
The increased mortality risk for prior children may come from shorter breastfeeding of 
girls to ensure that the mother can conceive again sooner, and with poor water
quality this shorter breastfeeding may result both in higher risk of dying and shorter
stature if the child survives \citep{Jayachandran2011,Jayachandran2017a}.
In other countries, focus has been mostly on the subsequent birth, but there is evidence 
in countries as diverse as Bangladesh, Brazil, El Salvadore, and Pakistan that
short intervals between births have adverse health effects
\citep{Cleland1984,Curtis1993,Davanzo2008,Gribble2009,Saha2013}.

The potential effects of spacing are not restricted to early outcomes, although
there is less evidence on longer-term effects and most comes from Western countries.
In the US, for example, a longer interval between births increases the older sibling's
test scores on the Peabody Individual Achievement Test, although there is no
statistically significant effect on the younger sibling \citep{Buckles2012}.
In line with this, close spacing appears to increase the risk of dropping out of
high school and lower the probability of attending a post-secondary school in both the
US and Sweden \citep{Powell1993,Pettersson-Lidbom2009}.
These results are, however, not uniformly supported, with recent results from Sweden 
showing no relationship between birth spacing, even if very short, and outcomes such as 
years of education completed, earnings, and unemployment \citep{Barclay2017}.

@@

\section{ }
